Component response rate variation underlies the stability of complex systems

### 2019-03-07 notebook/reviewer_comments_and_responses.tex

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\item
  Randomly generated matrices (\(M = XA\)) that are stable are more
  likely than not to have variation in X across the range of system
  sizes (S) that I simulated. This is (to my knowledge) a novel result
  and fully reprodible using the code underlying my simulations
  (\url{https://github.com/bradduthie/RandomMatrixStability}).
\item
  At high S, the probability that a system is stable becomes negligible;
  no such systems were found given S \textgreater{} 32. I now emphasise
  that my results apply specifically for the upper range of S
  \textgreater{} 10 for which stability is also non-negligible (i.e.,
  observed at least once in one million simulations). In this range,
  systems in which there is variation in X are indeed more stable than
  in systems in which no such variation exists. Further, this general
  pattern occurs for all simulated systems at the upper range of S,
  regardless of system connectance (C), interaction stengths
  (\(\sigma\)), or X and A element distributions (see Supplementary
  Information). In this manuscript, I am not interested in the effect of
  varying component response rates in systems that are too complex to be
  stable; my focus is entirely restricted to effects in the range where
  some stability can be reasonably expected. I explain this more
  carefully now in my Introduction and revise to state that my general
  conclusions may not apply as \(S \to \infty\).
\item
  Related to the above, I have, deliberately, not scaled interaction
  strengths (\(\sigma\)) by system size (\(S\)). My reason for doing
  this is because my manuscript is focused on the effect of component
  response rates as system complexity (\(\sigma \times \sqrt{SC}\))
  increases, not system size per se. Hence, S is increased while keeping
  all other parameter values constant so that I can evaluate the effect
  of \(Var(\gamma)\) on the stability of increasingly complex systems.
  This is fundamentally different from Gibbs et al. (2018), who scaled
  \(\sigma\) by \(S\) such that \(\sigma = 1/ \sqrt{S}\). I now state
  this explicitly in a new paragraph in the Discussion. I also show that
  my results are consistent with Gibbs et al. (2018) given the
  assumption that \(\sigma = 1/ \sqrt{S}\) in Supplementary Information.
\end{enumerate}

I hope these points adequately address Reviewer 3's concerns. There is
no contradiction between my results and those of Gibbs et al. (2018),
and I have revised my manuscript to more carefully explain why the two
works differ.

\subsection{Reviewer 3 specific comment
3}\label{reviewer-3-specific-comment-3}

This result is seemingly contradictory with the simulations reported in
the manuscript under review (Figs. 3 and 4). This is because the author
didn't scale the variance of \(A\) (\(\sigma^2\)) with \(1/S\), indeed
he maintained constant variance while increasing S. Increasing S without
scaling variances is misleading because eigenvalue distributions are not
comparable for different values of S. My feeling is that if he repeats
the figures 3 and 4 using this scaling for sigma\^{}2, then the
artifactual difference will disappear in the limit of large S, as Gibbs
et al found.

\subsection{Response to Reviewer 3 specific comment
3}\label{response-to-reviewer-3-specific-comment-3}

Reviewer 3 is of course correct that scaling the variance of A
(sigma\^{}2) with 1/S makes the difference between \(\gamma_{i} = 1\)
and \(Var(\gamma_{i})\) disappear. This was confirmed by simulation
during model development, but was not discussed in great detail in the
manuscript because it was not a focus of mine (this further highlights
the difference between my work and that of Gibbs et al. 2018). I now
explain this difference more, and I confirm Reviewer 3's feeling in the
Supplementary Information. I also explain the rational for note scaling
variances more clearly in the manuscript.

The reason for increasing values of S is to increase the total
complexity of the system (defined as \(\sigma \times \sqrt{SC}\)), and
to isolate and ultimately understand the effect of \(Var(\gamma_{i})\)
when S approaches a size at which a random system becomes too complex to
have a realistic probability of being stable (less than 1 in 1 million).
This is the reason for showing increasing system size (S) on the x-axis
of Figures 3 and 4. If I were to have scaled the variance of
off-diagonal elements of A (\(\sigma^{2}\)) with \(1/S\) as Reviewer 3
suggested (and as done in Gibbs et al. 2018), then the total complexity
of the system would not change, defeating the purpose of increasing S.
To my knowledge, there is also no a priori reason to expect systems to
have variances of interaction strengths (\(\sigma\)) that scale to their
size S, so forcing this to be the case was not justified.

Finally, while eigenvalue distributions are not \emph{identical} for
different values of S without scaling variances, it is not really
accurate to say that eigenvalue distributions are ``not comparable'';
eigenvalue distributions change with S in a predictable way that can be
compared when S is changed. It is well-established that the eigenvalues
for a random matrix M such as that used in my manuscript, and originally
by May (1972; Nature 238:413-414), will be uniformly distributed within
a circle on the complex plane within a radius of
\(\sigma\sqrt{SC} < d\), where \(-d\) is the mean of diagonal elements
of M (Tao and Tu 2010; Annals of Probability 38:2023-2065). Hence,
increasing S specifically increases the radius of this uniform
distribution, and does so predictably such that the probability that
real eigenvalues are negative (the criteria for system stability)
becomes increasingly low with increasing S. Multiplying the random
matrix A (which has the uniform distribution with a circle of radius
\(\sigma\sqrt{SC}\)) by the vector \(\gamma_{i}\) in my manuscript
resulted in an eigenvalue distribution that was more likely to be stable
at the upper range of S. I now clarify the difference between my result
and that of Gibbs et al. (2018) in this context in my revised
Discussion, and in a new section of the Supplementary Information.

\subsection{Reviewer 3 specific comment
4}\label{reviewer-3-specific-comment-4}

The author also states in this manuscript that stability can be largely
increased by `selecting' response rates so that linear stability is
forced to increase. This is an obvious technical result but (in the
light of the results by Gibbs et al.) it is also obvious that these
cases will form a set of null measure in the limit of large systems.

\subsection{Response to Reviewer 3 specific comment
4}\label{response-to-reviewer-3-specific-comment-4}

I showed that manipulating component response rates can, sometimes
greatly, increase the probability that a random system will be stable.
As with Reviewer 3 specific comment 3 (see above), I now clarify that my
focus is on the upper range of S at which large random systems can
reasonably be expected to be stable, not at the limit of
\(S \to \infty\) (as in Gibbs et al. 2018). Clarifying revisions have
been now made accordingly to the Introduction.

\section{Author References Cited}\label{author-references-cited}

Gibbs, T., Grilli, J., Rogers, T., \& Allesina, S. (2018). Effect of
population abundances on the stability of large random ecosystems.
Physical Review E, 98(2), 022410.

Gray, R. T., \& Robinson, P. A. (2009). Stability of random brain
networks with excitatory and inhibitory connections. Neurocomputing,
72(7--9), 1849--1858. \url{https://doi.org/10.1016/j.neucom.2008.06.001}

Gray, R. T., \& Robinson, P. A. (2008). Stability and synchronization of
random brain networks with a distribution of connection strengths.
Neurocomputing, 71(7--9), 1373--1387.
\url{https://doi.org/10.1016/j.neucom.2007.06.002}

Grilli, J., Adorisio, M., Suweis, S., Barabás, G., Banavar, J. R.,
Allesina, S., \& Maritan, A. (2017). Feasibility and coexistence of
large ecological communities. Nature Communications, 8.
\url{https://doi.org/10.1038/ncomms14389}

Hui, C., \& Richardson, D. M. (2018). How to invade an ecological
network. Trends in Ecology and Ecolution, xx, 1--11.
\url{https://doi.org/10.1016/j.tree.2018.11.003}

May, R. M. (1972). Will a large complex system be stable? Nature, 238,
413--414.

May, R. M. (1973). Qualitative stability in model ecosystems. Ecology,
54(3), 638--641. \url{https://doi.org/10.2307/1935352}

Patel, S., Cortez, M. H., \& Schreiber, S. J. (2018). Partitioning the
effects of eco-evolutionary feedbacks on community stability. American
Naturalist, 191(3), 1--29. \url{https://doi.org/10.1101/104505}

Haldane, A. G., \& May, R. M. (2011). Systemic risk in banking
ecosystems. Nature, 469(7330), 351--355.
\url{https://doi.org/10.1038/nature09659}

\end{document}
